*Prochlorococcus marinus* responses to light and oxygen

### 2024-03-05 Code/Manuscript_O2.Rmd
Adedd Bagby citation re CO2 and O2

---
title: "*Prochlorococcus marinus* responses to light and oxygen" 
author:
  - name: 'Mireille Savoie'
    institute: mta
    main: true
    email: msavoie@mta.ca
  - name: Aurora Mattison
    institute: [mta, ubc]
  - name: Julie Nadeau
    institute: mta
  - name: Laurel Genge
    institute: mta
  - name: Sylwia Śliwińska-Wilczewska
    institute: [mta, ug]
  - name: Maximilian Berthold
    institute: mta
  - name: Naaman M. Omar
    institute: mta
  - name: Ondřej Prášil
    institute: [mta, algatech]
    main: false
  - name: Amanda M. Cockshutt
    institute: [mta, sfx]
    main: false
  - name: Douglas A. Campbell
    institute: mta
    affil: 1
    main: false
    orcid: '0000-0001-8996-5463'
    email: 'dcampbel@mta.ca'
affiliation:
  - num: 1
    address: Departments of Biology and Chemistry and Biochemistry, Mount Allison University
institute:
  - mta: 'Department of Biology, Mount Allison University, 63B York St., Sackville NB, Canada, E4L 1G7.'
  - ubc: 'Department of Biology, University of British Columbia'
  - sfx: 'Department of Chemistry, St. Frances Xavier University'
  - algatech: 'Institute of Microbiology, Center Algatech, Laboratory of Photosynthesis, Novohradska 237, Trebon, CZ 37981, Czech Republic.'
  - ug: 'Institute of Oceanography, University of Gdansk, 46 Piłsudskiego St, Gdynia, Poland'
date: "`r format(Sys.Date())`"
output:
  bookdown::word_document2:
    code_folding: show
    reference_docx: KnitTemplate.docx
    keep_md: yes
    toc: FALSE
    toc_depth: 6
    fig_caption: yes
    pandoc_args: 
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
  bookdown::html_document2:
    code_folding: show
    keep_md: yes
    toc: TRUE
    toc_float: TRUE
    toc_depth: 6
    fig_caption: yes
    pandoc_args: 
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
always_allow_html: true
bibliography: [Manuscript_O2.bib, RPackages.bib, FaultyCitations.bib]
csl: plos-one.csl
editor_options: 
  markdown: 
    wrap: 72
---

@@

*Prochlorococcus marinus* MIT9313, a LLIV clade representative, shows potential to inhabit future warmer temperate zones year-round, as it grows under a 4 h photoperiod, expected in winter, or at light-attenuated depths. MIT9313 demonstrates an unexpected tolerance to higher light levels and cumulative diel PUR, but only under low oxygen conditions of 25 µM and 2.5 µM (Figure \@ref(fig:MIT9313PAR)), enabling MIT9313 to grow in OMZ, even at depths closer to the surface. MIT9313 carries a gene encoding (S)-2-hydroxy-acid oxidase [@cunaneCrystalStructureAnalysis2005],  with a K~M~ for [O~2~] of ~ 250 µM (Figure \@ref(fig:ProchlorococcusEnzymeKms)), which produces H~2~O~2~.  Growth at lower [O~2~] may protect MIT9313 from auto-intoxication from production of H~2~O~2~. We hypothesize that under 250 µM O~2~ and higher blue light, *P. marinus* MIT9313 suffers photoinhibition, resulting from the inactivation of PSII caused by the production of H~2~O~2~. This photoinhibition is compounded by the limited inducible repair for PSII, due to the absence of FtsH 1 and 2  expression in *P. marinus* MIT9313 [@bonisteelStrainSpecificDifferences2018].  MIT9313 shows remarkable ability to thrive under very low [O~2~], potentially allowing it to expand into broader ecological niches. These results are supported by Figure \@ref(fig:CladeProchloroPhotosynthDepthO2) showing PSI protein subunits annotated as derived from MIT9313 detected at depths \> 120 m, along with PSII subunits at depths from 50 m to 200 m in regions where O~2~ was 15 µM. The deep water environments typical for MIT9313 are relatively nutrient rich, and *Prochlorococcus* take up and metabolize various sugars [@gomez-baenaGlucoseUptakeIts2008; @munoz-marinGlucoseUptakeProchlorococcus2017; @munoz-marinProchlorococcusCanUse2013] and amino acids [@zubkovDepthRelatedAmino2004]. In future work we aim to test whether MIT9313 is using photosynthesis to drive CO~2~ fixation in low O~2~ environments, or whether PSII generation of O~2~ acts as an electron sink for respiration, using ATP for maintenance and to take up nutrients from the surroundings. Partensky *et al*. [@partenskyComparisonPhotosyntheticPerformances2018] found that in the low-light conditions found in the OMZ, MED4, SS120 and MIT9313 all became net O~2~ consumers, suggesting that the low light levels cause the respiratory chain to consume more O~2~ than the photosynthetic electron transport chain generates, thus contributing to maintenance of the low O~2~ environment.

>>>>

---
title: "*Prochlorococcus marinus* responses to light and oxygen" 
author:
  - name: 'Mireille Savoie'
    institute: mta
    main: true
    email: msavoie@mta.ca
  - name: Aurora Mattison
    institute: [mta, ubc]
  - name: Julie Nadeau
    institute: mta
  - name: Laurel Genge
    institute: mta
  - name: Sylwia Śliwińska-Wilczewska
    institute: [mta, ug]
  - name: Maximilian Berthold
    institute: mta
  - name: Naaman M. Omar
    institute: mta
  - name: Ondřej Prášil
    institute: [mta, algatech]
    main: false
  - name: Amanda M. Cockshutt
    institute: [mta, sfx]
    main: false
  - name: Douglas A. Campbell
    institute: mta
    affil: 1
    main: false
    orcid: '0000-0001-8996-5463'
    email: 'dcampbel@mta.ca'
affiliation:
  - num: 1
    address: Departments of Biology and Chemistry and Biochemistry, Mount Allison University
institute:
  - mta: 'Department of Biology, Mount Allison University, 63B York St., Sackville NB, Canada, E4L 1G7.'
  - ubc: 'Department of Biology, University of British Columbia'
  - sfx: 'Department of Chemistry, St. Frances Xavier University'
  - algatech: 'Institute of Microbiology, Center Algatech, Laboratory of Photosynthesis, Novohradska 237, Trebon, CZ 37981, Czech Republic.'
  - ug: 'Institute of Oceanography, University of Gdansk, 46 Piłsudskiego St, Gdynia, Poland'
date: "`r format(Sys.Date())`"
output:
  bookdown::html_document2:
    code_folding: show
    keep_md: yes
    toc: TRUE
    toc_float: TRUE
    toc_depth: 6
    fig_caption: yes
    pandoc_args: 
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
  bookdown::word_document2:
    code_folding: show
    reference_docx: KnitTemplate.docx
    keep_md: yes
    toc: FALSE
    toc_depth: 6
    fig_caption: yes
    pandoc_args: 
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
always_allow_html: true
bibliography: [Manuscript_O2.bib, RPackages.bib, FaultyCitations.bib]
csl: plos-one.csl
editor_options: 
  markdown: 
    wrap: 72
---

@@

*Prochlorococcus marinus* MIT9313, a LLIV clade representative, shows potential to inhabit future warmer temperate zones year-round, as it grows under a 4 h photoperiod, expected in winter, or at light-attenuated depths. MIT9313 demonstrates an unexpected tolerance to higher light levels and cumulative diel PUR, but only under low oxygen conditions of 25 µM and 2.5 µM (Figure \@ref(fig:MIT9313PAR)), enabling MIT9313 to grow in OMZ, even at depths closer to the surface. MIT9313 carries a gene encoding (S)-2-hydroxy-acid oxidase [@cunaneCrystalStructureAnalysis2005],  with a K~M~ for [O~2~] of ~ 250 µM (Figure \@ref(fig:ProchlorococcusEnzymeKms)), which produces H~2~O~2~.  Growth at lower [O~2~] may protect MIT9313 from auto-intoxication from production of H~2~O~2~. We hypothesize that under 250 µM O~2~ and higher blue light, *P. marinus* MIT9313 suffers photoinhibition, resulting from the inactivation of PSII caused by the production of H~2~O~2~. This photoinhibition is compounded by the limited inducible repair for PSII, due to the absence of FtsH 1 and 2  expression in *P. marinus* MIT9313 [@bonisteelStrainSpecificDifferences2018].  MIT9313 shows remarkable ability to thrive under very low [O~2~], potentially allowing it to expand into broader ecological niches. These results are supported by Figure \@ref(fig:CladeProchloroPhotosynthDepthO2) showing PSI protein subunits annotated as derived from MIT9313 detected at depths \> 120 m, along with PSII subunits at depths from 50 m to 200 m in regions where O~2~ was 15 µM. Bagby and Chisholm [@bagbyResponseProchlorococcusVarying2015] suggest that O~2~ has a protective role in *Prochlorococcus* under lower carbon dioxide environments when carbon fixation is limited. The deep water environments typical for MIT9313 are relatively nutrient rich, and *Prochlorococcus* take up and metabolize various sugars [@gomez-baenaGlucoseUptakeIts2008; @munoz-marinGlucoseUptakeProchlorococcus2017; @munoz-marinProchlorococcusCanUse2013] and amino acids [@zubkovDepthRelatedAmino2004].  In future work we aim to test whether MIT9313 is using photosynthesis to drive CO~2~ fixation in low O~2~ environments, or whether PSII generation of O~2~ acts as an electron sink for respiration, using ATP for maintenance and to take up nutrients from the surroundings. Partensky *et al*. [@partenskyComparisonPhotosyntheticPerformances2018] found that in the low-light conditions found in the OMZ, MED4, SS120 and MIT9313 all became net O~2~ consumers, suggesting that the low light levels cause the respiratory chain to consume more O~2~ than the photosynthetic electron transport chain generates, thus contributing to maintenance of the low O~2~ environment.
